---
title: PhD awarded
author: Phil Mike Jones
date: '2018-02-05'
slug: "2018-02-05-phd-awarded"
categories: [PhD]
tags: [PhD]
summary: "Today I received an email saying my PhD has been awarded, and my thesis has been published.
I am officially Dr Phil Jones!"
---

Today I received an email saying my PhD has been awarded, and my thesis has been published.
I am officially Dr Phil Jones!

```{r thesis-submitted, echo=FALSE}
blogdown::shortcode_html(
  "figure",
  src = "../../img/thesis-submitted.jpg",
  alt = "Submitting my thesis, September 2017",
  caption = "Submitting my thesis, September 2017")
```


## Manuscript and source code

The [manuscript and source code](http://etheses.whiterose.ac.uk/id/eprint/19283) are published on the White Rose consortium etheses repository.
You can also [download or fork the source code](https://github.com/philmikejones/thesis) from github.

```{r is-this-even-english, echo=FALSE}
blogdown::shortcode_html(
  "figure",
  src = "http://www.phdcomics.com/comics/archive/phd051807s.gif",
  alt = "'Piled Higher and Deeper' by Jorge Cham www.phdcomics.com",
  caption = "'Piled Higher and Deeper' by Jorge Cham www.phdcomics.com")
```

The manuscript is licensed under a [Creative Commons Attribution 2.0 UK: England & Wales](https://creativecommons.org/licenses/by/2.0/uk/) license and the source code is [GPL v2](https://www.gnu.org/licenses/old-licenses/gpl-2.0.en.html).
My hope is this makes it as easy as possible to build on this research.


## PhD statistics

A few statistics about the PhD, just out of curiosity...

### Dates

| Event                         | Date           |
|-------------------------------|----------------|
| Start                         | September 2013 |
| Submission                    | September 2017 |
| Viva                          | November 2017  |
| Submit corrections            | January 2018   |
| Thesis published; PhD awarded | February 2018  |

So, about 53 months from start to finish, not including a couple of leave of absences.


### Manuscript

| Item     | Number |
|----------|--------|
| Pages    | 271    |
| Chapters | 8      |
| Tables   | 27     |
| Figures  | 83     |


### Source code

| Item          | Number |
|---------------|--------|
| Lines of code | 23,053 |
| Commits       | 2,068  |
| Pull requests | 171    |

These don't include any work I did on the thesis before I converted it to bookdown in August 2016, or the work I did on [rakeR](https://philmikejones.github.io/rakeR/) (6,623 [LOC](https://en.wikipedia.org/wiki/Source_lines_of_code); 304 commits; 49 pull requests).
